---
title: "Qualitative Analysis of Weight Lifting Exercises"
author: "Alexander Klar"
date: "December 4, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

## Abstract

In this report we will investigate the dataset containing weight lifting exercises biceps curl and predict the quality of execution.
The original dataset is abailable here: http://groupware.les.inf.puc-rio.br/har.
The quality of the exercises is contained in the variable "classe" in the dataset: Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).
We will use the training set to create, train and crossvalidate a model to predict classe. 


## Data Preparation and Model Design

Download training data, create dataframe
```{r , cache=TRUE}
# download training data
trainingURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
trainingFile <- "pml-training.csv"
if(!exists(trainingFile)){
  download.file(trainingURL, destfile= trainingFile)
}
# create dataframe
trainingData <- read.csv(trainingFile, header = TRUE, na.strings = "NA")

library(caret)
# split into train and test
inTrain = createDataPartition(y=trainingData$classe, p = 0.7, list = FALSE)
training = trainingData[ inTrain,]
testing = trainingData[-inTrain,]

#modFitrf <- train(classe ~ ., data = training, method ="rf", prox=TRUE)
#modFitglm <- train(classe ~ ., data = training, method ="glm")
```



Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
